---
title: "Demonstration of R Markdown"
author: "JAS"
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## R Markdown Demonstration using Breast Imaging Data
Data Citation: This breast cancer database was obtained from the University of Wisconsin Hospitals, Madison from Dr. William H. Wolberg.
See Machine learning techniques to diagnose breast cancer from image-processed nuclear features of fine needle aspirates. Cancer Letters 1994; 77:163-171.

***

### Step 1: Load Packages
We will be using two packages in addition to base R.
Amelia is for missing data and will provide us with a plot of missingness.
Tidyverse includes a number of packages used for data management.

```{r packages}

#install.packages("tidyverse")

library(Amelia)
library(tidyverse)
library(here)
```
### Step 2: Load Data into R Environment and Perform Data Cleaning
#### Illustrates the following:
1. How to load a flat text file with no column headers
2. How to assign column names when none are provided
3. How to check variable types across the dataframe
4. How to recode missing indicators, change variable types and explore variable distributions
5. Create a quick plot to indicate missingness
6. Create a quick and dirty plot to compare features across outcome levels

```{r data_prepare}

#1. Replace file path to location of data file
bc.data =
  read_csv(here("./breast-cancer-wisconsin.data.txt"), col_names = FALSE) |>
  set_names(c("id", "clump_thickness", "uniformity_csize", "uniformity_cshape",
              "marg_adhesion", "single_ecell_size", "bare_nuclei",
              "b_chromatin", "normal_nucleoli", "mitoses", "outcome")) #2. Create vector of column names and append as column headers

#3. Check variable types: See that one is character even though all should be numeric. This is because a character "?" was used as the missing data indicator
str(bc.data)

#4. Replace the ? missing data indicator with NA, convert to numeric
bc.data_clean =
  bc.data |>
  mutate(
    bare_nuclei = if_else(bare_nuclei == "?", NA, bare_nuclei),
    bare_nuclei = as.numeric(bare_nuclei),
    outcome = factor(outcome, levels = c(2, 4), labels = c("Benign", "Malignant"))
)  #Convert outcome to factor and add appropriate labels

# Get quick summary of columns
summary(bc.data_clean)

#5. Quick visualization of missingness
missmap(bc.data_clean, main = "Missing values vs observed")

#6. Use ggplot to obtain a boxplot of feature by outcome
ggplot(bc.data_clean, aes(y=clump_thickness, x=outcome)) +  #Clump Thickness
  geom_boxplot()+
  theme_classic()

ggplot(bc.data_clean, aes(y=uniformity_csize, x=outcome)) +  #Uniformity size
  geom_boxplot()+
  theme_classic()

ggplot(bc.data_clean, aes(y=uniformity_cshape, x=outcome)) +  #Uniformity shape
  geom_boxplot()+
  theme_classic()

```

### Step 3: Construct logistic regression models to predict Malignancy

* Model 1: Include all features except ID
* Model 2: Include only clump thickness

```{r}
#Model 1: Using all features
bc.data_clean$id<-NULL #Setting ID variable to null

model.1 = 
  bc.data_clean |>
  glm(formula=outcome ~ ., family = binomial(link = 'logit'))

summary(model.1)

#Model 2: Specifying some features
model.2 =
  bc.data_clean |>
  glm(formula=outcome ~ clump_thickness, family = binomial(link = 'logit'))

summary(model.2)
```


