---
title: "Module 1 Exercise"
author: "Laura Robles-Torres"
date: "2025-06-28"
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Module 1 Exercise: 
These data represent clinical and demographic features of 286 female breast cancer patients in Yugoslavia. Women were followed from time of diagnosis onward for multiple years to determine if they had any recurrence of cancer. Data are in a flat-format, text file. No column headings (i.e. variables names) are provided. The features are a mix of ordinal and nominal variables. Data were obtained from the UCI Machine Learning Repository. This breast cancer domain was obtained from the University Medical Centre, Institute of Oncology, Ljubljana, Yugoslavia. Thanks go to M. Zwitter and M. Soklic for providing the data. Data can be obtained from the UCI Repository at this link:

http://archive.ics.uci.edu/ml/datasets/Breast+Cancer

***

### Step 1: Load Packages

```{r packages}

#install.packages("tidyverse")

library(Amelia)
library(tidyverse)
library(here)
library(dplyr)
library(forcats)
```

### Step 2: Load Data into R Environment and Perform Data Cleaning
1. Outcome Class: no-recurrence-events, recurrence-events 
2. Age at diagnosis: 10-19, 20-29, 30-39, 40-49, 50-59, 60-69, 70-79, 80-89, 90-99. 
3. When did menopause occur: lt40, ge40, premeno. 
4. tumor-size: 0-4, 5-9, 10-14, 15-19, 20-24, 25-29, 30-34, 35-39, 40-44, 45-49, 50-54, 55-59. 
5. # of auxiliary nodes that contain breast cancer: 0-2, 3-5, 6-8, 9-11, 12-14, 15-17, 18-20, 21-23, 24-26, 27-29, 30-32, 33-35, 36-39. 
6. Is the cancer in auxiliary nodes contained by the capsule of the lymph node: yes, no. 
7. Clinical degree of malignancy: 1, 2, 3. 
8. Location of primary tumor, breast: left, right. 
9. Location of tumor, within breast: left-up, left-low, right-up, right-low, central. 
10. Received radiation treatment: yes, no.

```{r data_prepare}

#1. Replace file path to location of data file
bc.data =
  read_csv(here("Module 1/breast-cancer.data.txt"), col_names = FALSE) |>
  set_names(c("outcome", "age_diagnosis", "menopause_status", "tumor_size",
              "cancerous_aux_nodes", "contained_by_capsule", "degree_malignancy",
              "location_breast", "location_within_breast", "radiation_tx")) #2. Create vector of column names and append as column headers

#3. Check variable types: See that one is character even though all should be numeric. This is because a character "?" was used as the missing data indicator
str(bc.data)

#4. Replace the ? missing data indicator with NA, convert to numeric
bc.data_clean =
  bc.data |>
  mutate(
    outcome = fct_relevel(outcome, "no-recurrence-events"),
    across(everything(), ~ if_else(. == "?", NA, .)),
    numnodes = case_when(
      cancerous_aux_nodes %in% c("0-2") ~ "0-2",
      cancerous_aux_nodes %in% c("3-5", "6-8") ~ "3-8",
      cancerous_aux_nodes %in% c("9-11", "12-14", "15-17", "18-20", "21-23",
                                 "24-26", "27-29", "30-32", "33-35", "36-39") ~ "More than 8"),
    numnodes = factor(numnodes, levels = c("0-2", "3-8", "More than 8"))
    ) |>
  select(-cancerous_aux_nodes)  #Convert outcome to factor and add appropriate labels


# Get quick summary of columns
summary(bc.data_clean)

#5. Quick visualization of missingness
missmap(bc.data_clean, main = "Missing values vs observed")

#6. Use ggplot to obtain a boxplot of feature by outcome
ggplot(bc.data_clean, aes(y=tumor_size, x=outcome)) + 
  geom_boxplot()+
  theme_classic()
```

### Step 3: Construct logistic regression models to predict Malignancy

* Model 1: Include all features except ID
* Model 2: Include only clump thickness

```{r}
#Model 1: Using all features
model.1 = 
  bc.data_clean |>
  glm(formula=outcome ~ ., family = binomial(link = 'logit'))

summary(model.1)

#Model 2: Specifying some features
model.2 =
  bc.data_clean |>
  glm(formula=outcome ~ degree_malignancy + radiation_tx + numnodes, family = binomial(link = 'logit'))

summary(model.2)
```



